---
title:  "US Inflation Rate Analysis"
output:
  html_document:
    df_print: paged
---

## Project Overview

This project analyzes the inflation rate in the United States from 2000 to 2021, focusing on data manipulation, matrix operations, and visualization in R. The data is sourced from a CSV file stored in Dropbox, and the project aims to compute annual inflation averages and visualize the trends over time.

## ==== First things first ====

Set working directory and call packages.

```{r}
# Check current working directory  
getwd()
# Set working directory

# Call required packages, ex. "readxl", "tidyverse"
if(!require(readxl)){install.packages("readxl")}
if(!require(tidyverse)){install.packages("tidyverse")}

```

## Analysis Steps

### Step 1: Import Data

**Objective**: Import U.S. inflation data from a CSV file in Dropbox and save it as `Inflation_US`. This file has no headers.

```{r}
# Import US inflation data from Dropbox
Inflation_US <- read.csv("https://www.dropbox.com/s/fffeqzozmo1w00w/Inflation_US.csv?raw=1", header = FALSE)

```

### Step 2: Check Object Type

**Objective**: Verify the data type of `Inflation_US`.

```{r}
# Check the data type of Inflation_US
class(Inflation_US)

```

### Step 3: Convert to Matrix

**Objective**: Create a matrix `Inflation_US_Matrix` with 22 rows, organized by row (byrow).

```{r}
# Convert data to matrix format
Inflation_US_Matrix <- matrix(data = as.vector(Inflation_US), nrow = 22, byrow = TRUE)

```

### Step 4: Set Row and Column Names

**Objective**: Assign row names (2000 to 2021) and column names (January to December) to `Inflation_US_Matrix`.

```{r}
# Set row names for years 2000 to 2021
rownames(Inflation_US_Matrix) <- 2000:2021

# Set column names using Thai month names
colnames(Inflation_US_Matrix) <- c("มกราคม", "กุมภาพันธ์", "มีนาคม", "เมษายน", 
                                   "พฤษภาคม", "มิถุนายน", "กรกฎาคม", "สิงหาคม", 
                                   "กันยายน", "ตุลาคม", "พฤศจิกายน", "ธันวาคม")

# Display the matrix
Inflation_US_Matrix

```

### Step 5: Calculate Annual Average Inflation

**Objective**: Calculate the average inflation rate per year and store it in `Avg_Inflation_US`.

```{r}
# Calculate row means for each year
Avg_Inflation_US <- rowMeans(Inflation_US_Matrix)

```

### Step 6: Add Annual Average Column to Matrix

**Objective**: Add a new column to `Inflation_US_Matrix` with the annual average inflation rate and label it as "Annual_Average".

```{r}
# Convert matrix to data frame and add average inflation column
Inflation_US_df <- as.data.frame(Inflation_US_Matrix)
Inflation_US_df$Annual_Average <- Avg_Inflation_US

```

### Step 7: Plot Annual Average Inflation

**Objective**: Create a line chart to visualize the annual average inflation rate from 2000 to 2021. The line is colored with RGB `#FF6666`, and has a thickness of 1 pt.

```{r}
# Prepare data frame for plotting
plot_data <- data.frame(Year = 2000:2021, Avg_Inflation_US = Avg_Inflation_US)

# Plot the annual average inflation rate
ggplot(plot_data, aes(x = Year, y = Avg_Inflation_US)) +
  geom_line(color = "#FF6666", size = 1) +
  labs(x = "Year", y = "Average Inflation Rate (%)") +
  theme_minimal()

```

Your name:Nutchanon

Student ID:

==== End-of-File ====
